Online {\IterComp}

### 2017-09-15 motivation.tex
taaaabs. sorry
Errrgh
changes to the result section; and other minor changes
results

    We performed offline iterative compilation on \texttt{susan\_c} to find the optimization settings that maximize the average performance
    over \FIXME{five} randomly chosen inputs from a set of 1,000 provided by the \textit{KDataSets} benchmark
	suite~\cite{chen10,chen12a}. The green histogram in Figure~\ref{fig:motivation-online} shows for each of the 1,0000 inputs the speedup versus \texttt{-O3}
    of that chosen optimization setting. The dark green line shows the average speedup for all those inputs. The
    \FIXME{five} training inputs are all in the top \FIXME{twelve} performing inputs. On the other hand, the chosen optimization does
    poorly on other inputs, often causing a slowdown and yielding only a slight speedup of 6\% on average.

    The blue bars in Figure~\ref{fig:motivation-online} show a different approach. Here the iterative compilation proceeds by evaluating
    each optimization on \FIXME{five} inputs. Each input is evaluated only once during the search, but the search is provided with oracle
    knowledge about the unoptimized runtime for the inputs, giving a perfect speedup metric. The environment of the search constantly changes
    in the same way as for online iterative compilation. The best optimization for this case never degrades performance compared to
    \texttt{-O3} and is on average 13\% better than the offline approach. The online case, considering truly representative inputs, leads
	to superior performance, if the speedup caused by optimizing can be estimated.

    \begin{figure}[t]
        \centering
        \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
        \caption{
            Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}.
            Each point represents averages over a subset of all inputs.
        }
        \label{fig:motivation-metric}
    \end{figure}

>>>>

    We performed offline iterative compilation on \texttt{susan\_c} to find the optimization settings that maximize the average performance
    over five randomly chosen inputs from a set of 1,000 provided by the \textit{KDataSets} benchmark suite~\cite{chen10,chen12a}.
    The green histogram in Figure~\ref{fig:motivation-online} shows for each of the 1,0000 inputs the speedup versus \texttt{-O3} of that
    chosen optimization setting. The dark green line shows the average speedup for all those inputs. The five training inputs are
    all among the top performing inputs. On the other hand, the chosen optimization does poorly on other inputs, often causing
    a slowdown and yielding only a slight speedup of 4\% on average.

    The blue bars in Figure~\ref{fig:motivation-online} show a different approach. Here the iterative compilation proceeds by evaluating
    each optimization on five inputs. Each input is evaluated only once during the search, but the search is provided with oracle
    knowledge about the unoptimized runtime for the inputs, giving a perfect speedup metric. The environment of the search constantly
    changes in the same way as for online iterative compilation. The best optimization for this case never degrades performance compared to
    \texttt{-O3} and is on average 13\% better than the offline approach. The online case, considering truly representative inputs, leads
    to superior performance, if the speedup caused by optimizing can be estimated.

    \begin{figure}[t]
        \centering
        \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
        \caption{
            Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}.
            Each point represents averages over a subset of all inputs.
            The IPC shows a correlation coefficient of $-0.14$.
        }
        \label{fig:motivation-metric}
    \end{figure}

### 2017-09-15 prof.tex
results

Graph theory shows that we can calculate precisely the frequencies of all edges between basic blocks if we choose a spanning tree of the
function CFG and instrument the edges not in the tree. To make the instrumentation probe placement optimal, we choose the maximum spanning
tree, with the edge weights representing edge frequency estimates. By doing so, we leave as many high frequency edges as possible in the
maximum spanning tree and they will not be instrumented, allowing us to minimize the instrumentation overhead.

@@

\paragraph{Populating the edge flows:}
Intuitively, if all the edge flows are known for the complement of a spanning tree then at any leaf of the spanning tree there is only one
unknown edge flow. This unknown edge flow can be calculated by Kirchhoff's first law\FIXME{\cite{}}. This process repeats as a bottom-up
propagation until all the unknown edge flows have been calculated. This algorithm can be formally defined as a post-order traversal on the
spanning tree. Let $G$ be the CFG. This algorithm first initializes the edge flows $D_{(u,v)}$, for each edge $(u,v)$ in the CFG, as
follows:
\[
D_{(u,v)} \gets
\begin{cases}
    P_{(u,v)} & \quad \text{if edge $(u,v)$ has a probe $P_{(u,v)}$}\\
    0       & \quad \text{otherwise}
\end{cases}
\]
Afterwards, for each vertex $u\in G$, in a post-order traversal of the spanning tree,
we can use the Kirchhoff's first law and apply the following operations
in a symbolic fashion:
\[
\Sigma^+_u = \sum_{v\in N^+(u)} D_{(u,v)}
\]
\[
\Sigma^-_u = \sum_{v\in N^-(u)} D_{(v,u)}
\]
\[
\forall v\in N^+(u):  D_{(u,v)} \gets
\begin{cases}
    D_{(u,v)} & \quad \text{if $D_{(u,v)}\neq 0$}\\
    \Sigma^-_u - \Sigma^+_u       & \quad \text{otherwise}
\end{cases}
\]
\[
\forall v\in N^-(u):  D_{(v,u)} \gets
\begin{cases}
    D_{(v,u)} & \quad \text{if $D_{(v,u)}\neq 0$}\\
    \Sigma^+_u - \Sigma^-_u       & \quad \text{otherwise}
\end{cases}
\]

>>>>

Graph theory shows that we can calculate precisely the frequencies of all edges between basic blocks if we choose a spanning tree of the
function CFG and instrument the edges in the complement of the spanning tree~\cite{nahapetian73,forman81}.
To make the instrumentation probe placement optimal, we choose the maximum spanning tree,
with the edge weights representing edge frequency estimates. By doing so, we leave as many high frequency edges as possible in the
maximum spanning tree and they will not be instrumented, allowing us to minimize the instrumentation overhead~\cite{forman81,ball94}.

@@

\paragraph{Populating the edge flows:}
Intuitively, if all the edge flows are known for the complement of a spanning tree then at any leaf of the spanning tree there is only one
unknown edge flow. This unknown edge flow can be calculated by Kirchhoff's first law\cite{knuth73,ball94}. This process repeats as a bottom-up
propagation until all the unknown edge flows have been calculated. This algorithm can be formally defined as a post-order traversal on the
spanning tree. Let $G$ be the CFG. This algorithm first initializes the edge flows $D_{(u,v)}$, for each edge $(u,v)$ in the CFG, as
follows:
\[
D_{(u,v)} \gets
\begin{cases}
    P_{(u,v)} & \quad \text{if edge $(u,v)$ has a probe $P_{(u,v)}$}\\
    0       & \quad \text{otherwise}
\end{cases}
\]
Afterwards, for each vertex $u\in G$, in a post-order traversal of the spanning tree,
we can use the Kirchhoff's first law and apply the following operations
in a symbolic fashion:
\[
\Sigma^+_u = \sum_{v\in N^+(u)} D_{(u,v)}
\]
\[
\Sigma^-_u = \sum_{v\in N^-(u)} D_{(v,u)}
\]
\[
\forall v\in N^+(u):  D_{(u,v)} \gets
\begin{cases}
    D_{(u,v)} & \quad \text{if $D_{(u,v)}\neq 0$}\\
    \Sigma^-_u - \Sigma^+_u       & \quad \text{otherwise}
\end{cases}
\]
\[
\forall v\in N^-(u):  D_{(v,u)} \gets
\begin{cases}
    D_{(v,u)} & \quad \text{if $D_{(v,u)}\neq 0$}\\
    \Sigma^+_u - \Sigma^-_u       & \quad \text{otherwise}
\end{cases}
\]

### 2017-09-15 results.tex
results

A few of the benchmark present abnormal regressions in their execution time.
This increase in the profiling overhead is counter-intuitive because the
relaxation algorithms only reduces the number of instrumented probes,
without changing their placement.
We suggest some reasons for this abnormal regressions.
By analysing these cases, we noticed that several of the frequently executed
probes could not be removed by the relaxation.
Moreover, another important aspect to point out is that we also observed
differences among the final CFGs after applying the \texttt{-O3} optimisation.
Although relaxation does not directly alter the CFG, it is worth mentioning that
many optimisation passes make use of LLVM's built-in target-based cost model,
which means that adding or removing instructions from basic blocks may affect
decisions during transformations of the code.
For example, some of the optimizations that make use of this built-in cost model
and also alter the CFG are: loop-unrolling, function inlining, and
simplifications of the CFG.
These changes in the program's CFGs may affect other compiler optimisations and
also possibly in hardware performance, for example affecting the instruction
caching.

@@

In order to evaluate the quality of the final optimization sequences selected by
each configuration of the online {\itercomp}, we compare their speedup over the
{\flagstype -O3} optimization across all the 1,000 inputs of the benchmark
being optimized.
%When measuring the wall-clock time for each input, to reduce noise, we execute
%the same input until we have a statistically sound measurement, i.e. we execute
%until we have an interval no larger than 1\% with 99\% confidence.
Figure~\ref{fig:speedups} shows these average speedups for all test benchmarks.
This figure shows that the best optimization sequence selected with the Oracle-PP
is very close to the performance of the best optimization sequence selected with
the Oracle-RM, where Oracle-PP achieves on average about 80\% of the performance
improved obtained by Oracle-RM.
This result is important for demonstrating that a work-based metric has the
potential to produce good results in a real-world online scenario.

Moreover, the evaluation also indicates that the use of relaxation algorithms
does not degrade the optimization search, it might even benefit the {\itercomp}
as it tends to insert smaller interferences in the performance measurement.
%Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.

%\begin{figure*}[h]
%    \centering
%    \includegraphics[width=\textwidth]{figs/ipc-vs-work.pdf}
%    \caption{Comparison between {\itercomp} using IPC or the WP metric.}
%    \label{fig:ipc-vs-work}
%\end{figure*}

%Figure~\ref{fig:ipc-vs-work} presents a comparison between IPC and the WP metric regarding their correlation with the actual speedup over \verb|-O0| observed during {\itercomp}.
%Each point in the figure corresponds to each metric averaged over input windows collected during multiple executions of the {\itercomp}.
%The vertical lines correspond to the optimization sequence selected by the {\itercomp} search, which either represents the highest IPC or the highest value of WP.
%In all cases, the optimization sequence selected based on the WP is much closer to the best speedup than the optimization sequence selected based on IPC, which can be attributed to the fact that the WP has a stronger correlation with the speedup.
%In all cases, IPC shows little to no correlation with the speedup, which corroborates the argument given in Section~\ref{sec:ipc-vs-work-metric}.

%For some cases, namely for the tiff-related benchmarks, the WP shows a worse correlation with speedup.
%We believe these cases could be improved by improving the cost model used to compute the work metric.
%However, even in these cases, the optimization sequences selected based on the WP outperform those selected based on the IPC metric.

%\textbf{ToDo: Include results using Genetic Algorithm search.}

>>>>

A few of the benchmark present abnormal regressions in their execution time.
This increase in the profiling overhead is counter-intuitive because the
relaxation algorithms only reduces the number of instrumented probes,
without changing their placement.
We suggest some reasons for this abnormal regressions.
By analysing these cases, we noticed that several of the frequently executed
probes could not be removed by the relaxation.
Moreover, another important aspect to point out is that we also observed
differences among the final CFGs after applying the \texttt{-O3} optimization.
Although relaxation does not directly alter the CFG, it is worth mentioning that
many optimization passes make use of LLVM's built-in target-based cost model,
which means that adding or removing instructions from basic blocks may affect
decisions during transformations of the code.
For example, some of the optimizations that make use of this built-in cost model
and also alter the CFG are: loop-unrolling, function inlining, and
simplifications of the CFG.
These changes in the program's CFGs may affect other compiler optimizations and
also possibly in hardware performance, for example affecting the instruction
caching.

@@

In order to evaluate the quality of the final optimization sequences selected by
each configuration of the online {\itercomp}, we compare their speedup over the
\texttt{-O3} optimization across all the 1,000 inputs of the benchmark being
optimized.
None of the configurations degrades the performance over \texttt{-O3} (with
statistical significance).

%When measuring the wall-clock time for each input, to reduce noise, we execute
%the same input until we have a statistically sound measurement, i.e. we execute
%until we have an interval no larger than 1\% with 99\% confidence.
Figure~\ref{fig:speedups} shows these average speedups for all test benchmarks.
This figure shows that the best optimization sequence selected with the Oracle-PP
is very close to the performance of the best optimization sequence selected with
the Oracle-RM, where Oracle-PP achieves on average about 80\% of the performance
improvement obtained by the Oracle-RM.
This result is important for demonstrating that our work efficiency metric has
the potential to produce good results in real-world online scenarios.

The Real-OP achieves 4\% improvement on average, which represents 45\% of the
performance improvement obtained by the Oracle-RM.
This difference to Oracle-PP is explained by the real (intrusive) profiling used
in the Real-OP configuration.
The profiling affects the search in two key ways:
\textit{(i.)} the overhead incurred by the profiling affects the execution time,
and therefore also affects the work efficiency metric,
\textit{(ii.)} the instrumentation code may affect many of the optimizations, e.g.,
due to the use of a global variable or in decisions taken based on a cost-model
for the instructions.
However, even in the Real-OP configuration, which has a highly intrusive instrumentation,
it is still possible to obtain performance improvements in a realistic online
{\itercomp}.

Finally, the evaluation also indicates that the use of relaxation algorithms
is beneficial not only for its reduction in overhead, which is directly observed
by the user, by it also tends to improve the quality of the search.
This improvement has to do with the aforementioned effects of an intrusive
instrumentation on the search.
The relaxation has the potential to reduce both effects of the profiling.
Both the Real-R5 and Real-WPO-R5 achieve about 60\% of the
performance improvement obtained by the Oracle-RM.

### 2017-09-15 work.tex
changes to the result section; and other minor changes
results

    \[
       P = \nicefrac{\Delta W}{\Delta t}
    \]

@@

    Figure~\ref{fig:cost-model} shows the correlation of the model with runtime on instances of test benchmarks. The correlation
    coefficient is \FIXME{XXX}, and the mean absolute percentage error (MAPE) is \FIXME{XXX}. Figure~\ref{fig:motivation-work-efficiency}
    shows a similar plot to Figure~\FIXME{\ref{fig:???}}, of work efficiency versus speedup for 500 different binaries of \texttt{susan\_c}. Here the new metric correlates much better than either performance or IPC.

@@

    \begin{figure}[t]
        \centering
        \includegraphics[width=0.5\textwidth]{figs/motivation-work-efficiency.pdf}
        \caption{Relationship between work efficiency vs speedup for 500 different binaries of \texttt{susan\_c}.
                Each point represents averages over a subset of all inputs.}
        \label{fig:motivation-work-efficiency}
    \end{figure}

>>>>

    \[
       \textrm{Work Efficiency} = \frac{\Delta W}{\Delta t}
    \]

@@

    Figure~\ref{fig:cost-model} shows the correlation of the model with runtime on instances of test benchmarks. The correlation
    coefficient is $0.99$, and the mean absolute percentage error (MAPE) is about 24\%.
    Figure~\ref{fig:motivation-work-efficiency} shows a similar plot to Figure~\FIXME{\ref{fig:motivation-metric}},
    of work efficiency versus speedup for 500 different binaries of \texttt{susan\_c}.
    Here the new metric correlates much better than either execution time or IPC,
    exhibiting a correlation coefficient of $0.93$ with speedup.

@@

    \begin{figure}[t]
        \centering
        \includegraphics[width=0.5\textwidth]{figs/motivation-work-efficiency.pdf}
        \caption{Relationship between work efficiency vs speedup for 500 different binaries of \texttt{susan\_c}.
                Each point represents averages over a subset of all inputs.
                Correlation coefficient of $0.93$.}
        \label{fig:motivation-work-efficiency}
    \end{figure}
